---
title: "Some functions and definitions"
output: 
  html_notebook: 
    code_folding: hide
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    theme: readable
    toc: yes
  bookdown::word_document2:
    toc: no
    latex_engine: xelatex
    header-includes:
      - \usepackage{setspace}
      - \doublespacing
    fontsize: 12pt
    font: Times New Roman
    reference_docx: wordtemplate.docx
bibliography: reference.bib
csl: GCA.csl
editor_options: 
  markdown: 
    wrap: 72
---

## Time scales for Lutao fluid samples

```{r Lutao-fluid, eval=FALSE, results="none", echo=FALSE}
theme.chen2 <- theme(axis.title = element_text(size=18),
                     axis.text = element_text(size=14),
                     legend.text = element_text(size=12),
                     legend.title = element_text(size=14))

theme.chen1 <-  theme(
  axis.title = element_text(size = 18),
  axis.text = element_text(size = 14),
  legend.text = element_text(size = 12),
  legend.title = element_text(size = 14),
  axis.ticks.length = unit(0.15,"cm"),
  # axis.text.x.bottom = element_text(margin=unit(c(0.5,0.5,0.5,0.5), "cm")),
  # axis.text.y.left = element_text(margin=unit(c(0.5,0.5,0.5,0.5), "cm")),
  # axis.text.x.top = element_text(margin=unit(c(0.5,0.5,0.5,0.5), "cm")),
  # axis.text.y.right = element_text(margin=unit(c(0.5,0.5,0.5,0.5), "cm")),
  axis.line = element_line(size = 1),
  # axis.title.x.bottom = element_text(margin=unit(c(-.4,-.1,-.1,-.1), "cm")),
  # axis.title.y.left = element_text(margin=unit(c(-.1,-.4,-.1,-.1), "cm")),
  panel.background = element_blank()
)

insert_minor <- 
  function(major_labs, n_minor,n_breaks) {
    labs <- c( sapply( major_labs, 
                       function(x) c(x, rep("", n_breaks) ) ) )
                              labs[1:(length(labs)-n_minor)]
                              }


```

## (P,$\rho$, T) properties of standard seawater. by Safarov et al., 2012,2013

absolute salinity SA=0-55.529 g kg<sup>-1</sup>, T = 273.15 - 468.15 K,
P up to 140 Mpa (1400 bar).

$$P=A(\rho)^2+B(\rho)^8+C(\rho)^{12}$$ $\rho$ in g cm<sup>-3</sup>, P in
MPa, T in K

$$A=\Sigma^5_{i=1}*T^i*\Sigma^2_{j=0}*a_{ij}*S_A^j$$
$$B=\Sigma^4_{i=0}*T^i*\Sigma^2_{j=0}*b_{ij}*S_A^j$$
$$C=\Sigma^4_{i=0}*T^i*\Sigma^2_{j=0}*c_{ij}*S_A^j$$ **The coefficients
for aij, bij, and cij are shown in the Table 2 of Safarov et al.,
2012**.

```{r eval=FALSE}
# seawater density as a function of T, P, and salinity
# Safarov et al., 2009, 2012, 2013
# T in oC and P in bar
Seawater_Density <- function (T, P, S) {
  TK <- T+273.15
  P_mpa <- P/10
  
  a10 <- 16.8260119036
  a11 <- -0.0141313351238
  a12 <- 0.247406490575/1000
  a20 <- -0.174551209401
  a21 <- 0.336290751342/10^4
  a22 <- -0.102404029138/10^5
  a30 <- 0.590549251552/10^3
  a31 <- 0
  a32 <- 0
  a40 <- -0.843286627505/10^6
  a41 <- 0
  a42 <- 0.482122273652/10^11
  a50 <- 0.444848916045/10^9
  a51 <- -0.346870313329/10^13
  a52 <- -0.534550469142/10^14
  A1 <- TK*(a10*S^0+a11*S+a12*S^2)
  A2 <- TK^2*(a20*S^0+a21*S+a22*S^2)
  A3 <- TK^3*(a30*S^0+a31*S+a32*S^2)
  A4 <- TK^4*(a40*S^0+a41*S+a42*S^2)
  A5 <- TK^5*(a50*S^0+a51*S+a52*S^2)
  A <- A1+A2+A3+A4+A5
  
  b00 <- 1780.94028549
  b01 <- -8.44625953277
  b02 <- -0.0203022580799
  b10 <- -23.9619103076
  b11 <- 0.0616367739239
  b12 <- 0
  b20 <- 0.123588063860
  b21 <- -0.115367580107/10^3
  b22 <- 0
  b30 <- -0.230808294277/10^3
  b31 <- 0
  b32 <- 0
  b40 <- 0.142119761663/10^6
  b41 <- 0
  b42 <- 0.534482054293/10^12
  B1 <- (b00+b01*S+b02*S^2)
  B2 <- TK*(b10+b11*S+b12*S^2)
  B3 <- TK^2*(b20+b21*S+b22*S^2)
  B4 <- TK^3*(b30+b31*S+b32*S^2)
  B5 <- TK^4*(b40+b41*S+b42*S^2)
  B <- B1+B2+B3+B4+B5
  
  c00 <- -1197.00357946
  c01 <- 8.48513299409
  c02 <- 0
  c10 <- 11.7622451237
  c11 <- -0.0571254359702
  c12 <- 0
  c20 <- -0.0402338524280
  c21 <- 0.958997239703/10^4
  c22 <- 0.185255385513/10^6
  c30 <- 0.418262515254/10^4
  c31 <- 0
  c32 <- 0
  c40 <- 0
  c41 <- 0
  c42 <- -0.986850131470/10^12
  C1 <- (c00+c01*S+c02*S^2)
  C2 <- TK*(c10+c11*S+c12*S^2)
  C3 <- TK^2*(c20+c21*S+c22*S^2)
  C4 <- TK^3*(c30+c31*S+c32*S^2)
  C5 <- TK^4*(c40+c41*S+c42*S^2)
  C <- C1+C2+C3+C4+C5
  
  # P_mpa = A*\rho^2+B*\rho^8+C*\rho^12
  # using polyroot to find the density of seawater, d (g cm-3)
  
  densityfunction <- function (x) {
    A*x^2+B*x^8+C*x^12-P_mpa
  }
  
  density <- uniroot(densityfunction,lower = 0.8,upper = 1.2)$root
  
  return(density)
}

```

## critical curve of seawater: Driesner and Heinrich, 2007

**Equation 5a** for temperatures below critical temperature of pure
water (373.976 <sup>o</sup>C).

`This equation is used for pure water, however, at temperature of below critical point of pure water, the critical curves of pure water and seawater are almost identical (Fig. 1)`
$$P_{crit}=P^{H2O}_{crit}+\Sigma^7_{n=1}c_n(T^{H2O}_{crit}-T)^{c_nA}$$
P<sub>crit</sub>=220.54915 bar The values for the constants are shown in
Table 4.

```{r eval=FALSE}
critical_pressure <- function (T, type) {
  criticalPressureH2O <- 220.54915
  criticalTempH2O <- 373.976
  c1 <- -2.36
  c2 <- 1.28534/10
  c3 <- -2.3707/100
  c4 <- 3.20089/1000
  c5 <- -1.38917/10^4
  c6 <- 1.02789/10^7
  c7 <- -4.8376/10^11
  c1a <- 1
  c2a <- 1.5
  c3a <- 2
  c4a <- 2.5
  c5a <- 3
  c6a <- 4
  c7a <- 5
  if (type=="full") {
    P <- criticalPressureH2O+c1*(criticalTempH2O-T)^c1a+c2*(criticalTempH2O-T)^c2a+c3*(criticalTempH2O-T)^c3a+c4*(criticalTempH2O-T)^c4a+c5*(criticalTempH2O-T)^c5a+c6*(criticalTempH2O-T)^c6a+c7*(criticalTempH2O-T)^c7a
  } else if(type=="saturated") {
    if (T>=100) {
      P <- criticalPressureH2O+c1*(criticalTempH2O-T)^c1a+c2*(criticalTempH2O-T)^c2a+c3*(criticalTempH2O-T)^c3a+c4*(criticalTempH2O-T)^c4a+c5*(criticalTempH2O-T)^c5a+c6*(criticalTempH2O-T)^c6a+c7*(criticalTempH2O-T)^c7a
    } else P <- 1
  } else print("The type should be either full or saturated!")
  return(P)
}

```

## Quartz solubility

### Von Damm et al., 1991

$$lnM=a+bln\rho+(C+dT^2)/T+eP/T$$ where M is the molarity of dissolved
SiO~2~, $\rho$ is the density of the solution (g cm^-3^), P is the
pressure (bar), T is temperature in *K*.

a=-2.32888, b = 1.79547, c=-2263.62, d=0.00407350, e=0.0398808.

This equation is proposed to be valid at temperature range from 45
<sup>o</sup>C to 900 <sup>o</sup>C and pressure up to 9860 bars.

### Gunnarsson and Arnorsson, 2000

$$logK_{am.silica}=-8.476-485.24 \times T^{-1} - 2.268 \times 10^{-6} \times T^2+3.068\times logT$$
$$logK_{quartz}=-34.188+197.47 \times T^{-1} - 5.851 \times 10^{-6} \times T^2+12.245\times logT$$
T in K. Valid at 0 - 350 ^o^C at 1 bar below 100 ^o^C and P~sat~ at
higher temperatures. SiO~2~ in ppm.

```{r quartz-solubility, eval=FALSE, results="none"}
# SiO2 equilibrate with quartz as a function pf T and P, Von Damm et al., 1991
# Laboratory study, valid at 45 - 900 oC
#density as a function of Temperature, pressure, and density, Von Damm et al., 1991
Quartz_sol_vonDamm1991 <- function (T, S) {
  
  # seawater density, valid at 273.15 - 468.15 K
  
  P <- critical_pressure(T=T,type = "saturated")
  d <- Seawater_Density(T, P, S)

  TK <- T+273.15
  ln_Sol <- -2.32888 + 1.79547 * log(d, base = exp(1)) + (-2263.62 + 0.00407350 * TK * TK) / TK + 0.0398808 * P / TK
  Sol_mM <- 1000 * exp(ln_Sol)
  return(Sol_mM)
}

Quartz_sol_Gunnarsson2000 <- function (T) {
  TK <- T+273.15
  logK_ppm <- -34.188+197.47/TK-5.851*10^-6*TK^2+12.245*log10(TK)
  SiO2_mM <- (10^logK_ppm)*1000
  return (SiO2_mM)
}


```

## Anhydrite dissolution by Arnorsson et al., 1982

$$CaSO_4(s)=Ca^{2+}+SO_4^{2-}$$ $$log10(K)=3.94-677/T-12.39*T/1000$$
This equation is valid at 0-370 <sup>o</sup>C.

```{r anhydrite, eval=FALSE, results="none"}
# equilibrium constant of anhydrite dissolution
# valid at 0-370 oC
Anhydrite_Dissolution_constant <- function (T) {
  TK <- T+273.15
  logK <- 3.94-677/TK-12.39/1000*TK
  return(logK)
}

```

## Na-K geothermometer by Giggenbach 1988

$$log10(K/Na)=1.75-1390/T$$ T in K and concentrations in mg/kg.

## Henry's law

There are several values to express Henry's law constant.
$K^{px}_H=\frac {p}{x}$ with a unit of atm. The higher of this value,
the lower proportion of gas dissolution. $H^{cc}=\frac{c_{aq}}{c_{gas}}$
which is dimensionless. The higher of this value, the higher gas
dissolution degree.

**Argon**

According to Krause and Benson (1989).
$$T^{*2}ln(K_H)=T^{*2}A_0+A_1(1-T^*)^{1/3}+A_2(1-T^*)^{2/3}$$
$T^*\equiv T/T_{cl}$, T~cl~ is the critical point of water (647 K).
A~0~=4.289125, A~1~=19.225988, A~2~=-21.603721.

**Warning: This equation is derived from experiments at 0 - 60 ^o^C.**
But t fits the high temperature data by other researchers.

**Nitrogen**

$$K^{px}_H=\frac{1}{10^{-pB}/55.55}$$
$$pB=9.755-\frac{1121}{T}-0.009169*T$$ where T is temperature in K.
**This equation is valid at 0 -350 ^o^C**.

Therefore, the Ar and N~2~ that dissolved in the fluid could be
expressed as: $$Ar_{aq}\sim Ar_{air}/K_{Ar}$$
$$N2_{aq}\sim N2_{air}/K_{N2}$$
$$(\frac{N_2}{Ar})_{aq}=(\frac{N_2}{Ar})_{air}*(\frac{K_{Ar}}{K_{N2}})$$
**The code**

```{r}
Henry_Ar <- function (T) {
  Tcl <- 647
  T0 <- (T+273)/Tcl
  a0 <- 4.289125
  a1 <- 19.225988
  a2 <- -21.603721
  ln_KH <- a0+a1*(1-T0)^(1/3)/T0^2+a2*(1-T0)^(2/3)/T0^2
  KH <- exp(ln_KH)
  return (KH)
}

Henry_N2 <- function (T) {
  pB <- 9.755-1121/(T+273)-0.009169*(T+273)
  Hk <- 10^(-pB)/55.55
  KH <- 1/Hk
  return (KH)
}

N2_Ar <- function (T) {
  Ar_air <- 9.32
  N2_air <- 781
  ratio <- N2_air/Ar_air*Ar_henry(T)/N2_henry(T)
  return (ratio)
}

# The theoretical line of Ar
ggplot(data = data.frame(x=c(0,300)),aes(x=x))+
  stat_function(fun = Ar_henry,geom = "line",color="tomato2",size=1)+
  xlab("Temperature (degree C)")+
  ylab(expression(paste(K[H])))

# The theoretical line of N2
ggplot(data = data.frame(x=c(0,300)),aes(x=x))+
  stat_function(fun = N2_henry,geom = "line",color="tomato2",size=1)+
  xlab("Temperature (degree C)")+
  ylab(expression(paste(K[H])))

# The theoretical line of N2/Ar
ggplot(data = data.frame(x=c(0,300)),aes(x=x))+
  stat_function(fun = N2_Ar,geom = "line",color="tomato2",size=1)+
  xlab("Temperature (degree C)")+
  ylab(expression(paste(K[H])))
```

## Horita, 2001

### input temperature, return $\Delta_{CO_2-CH_4}$

```{r dpi=300, echo=TRUE, eval=FALSE}
Horita_T <- function (T) { #input temperature in K, return dCO2-CH4
  
   dCO2_CH4 <- 0.16+11.754*10^6/(T+273)^2-2.3655*10^9/(T+273)^3+0.2054*10^12/(T+273)^4 #the delta (CO2-CH4) if isotopic equilibrium achieved, Horita, 2001 
   dd <- data.frame(Temp=T,delta=dCO2_CH4) #made a dataframe for plotting
   return (c("The fractionation between CO2 and CH4 is", dCO2_CH4))  

}

```

### input $\Delta_{CO_2-CH_4}$, return temperature

```{r dpi=300, echo=TRUE, eval=FALSE}
library(tidyverse)
library(latex2exp)

Horita <- function (dC) {                     #input dCO2-CH4, return temperature in Celcius 
  df <- function (temperature) {              #input temperature in K, return dCO2-CH4
    dCO2_CH4 <- 0.16+11.754*10^6/(temperature+273)^2
    -2.3655*10^9/(temperature+273)^3
    +0.2054*10^12/(temperature+273)^4         #the delta (CO2-CH4) if isotopic equilibrium
                                              #achieved, Horita, 2001 
    dd <- data.frame(Temp=temperature,delta=dCO2_CH4)     #made a dataframe for plotting
    return (dd) 
    }
  
  temperature <- seq(1,1000,1)
  df <- df(temperature)
  
  if (df$delta[1]<dC) {
    return("Warning: the input is larger than the maximum theoretical value!")
  } else if (df$delta[1000]>dC) {
    return("Warning: the input is smaller than the minimum theoretical value!")
  } else {
    
  n <- which.min(abs(df$delta-dC))             #return the nearest delta value at temperature
  message <- c("The temperature at ΔCO2-CH4 of",dC," is ",n," degree C") #return the message
  
  plot <- df %>%                              #draw a plot to show the calculation
    ggplot(aes(x=Temp,y=delta))+
    geom_path(color="springgreen3",
              size=1)+
    annotate("segment",x = c(0,n),
             xend = c(n,n),
             y = c(dC,0),
             yend = c(dC,dC),
             lty="dashed")+
    xlab(expression(paste("Temperature (",phantom()^o,"C)")))+
    ylab(TeX('$\\Delta_{CO_2-CH_4}$'))+
    theme.chen2
  
    return (list(message,plot))
  
  }
  
}

```

## try to discover the endmember \$\\delta\$^13^C of CO~2~

-   at 90-100 ^o^C, fraction of residual CO2 after dissolution will be
    85.9 % - 83.6% (Chen et al., 2018, equation 10)

-   fractionation between gaseous CO2 and dissolved carbonate:
    $$10^{3}ln\alpha(CO_2-CH_4) = 0.16 +11.754*(10^6/(T+273)^2)-2.3655*(10^9/(T+273)^3)+0.2054*(10^12/(T+273)^4)$$

$$\Delta CO_2-HCO_{3}^- = -(9.866\pm0.23)*10^3/T+(24.12\pm0.78)‰$$ (Mook
et al., 1974)

$$\epsilon _{aq-g} = -(0.0049\pm0.003)*T(^oC) - (1.31\pm0.06)‰$$ (Zhang
et al., 1995)

## CO2 disolution

According to Henry's law `(Sander, 2015)`, the dissolution of gas in an
ideal state can be expressed as: $$H^{cc} = C_{aq}/C_{gas}$$

where C~aq~ and C~gas~ are the gas dissolved in the aqueous phase and
remained in the gas phase, respectively. *H^cc^* is the dimensionless
Henry's law constants, which changes with temperature:
$$H^{cc}=H_{\theta}^{cc}*e[-\Delta_{sol}H/R(1/T-1/T\theta)]$$ where
$H^{cc}_{\theta}$ is the Henry's law constant at *T*$\theta$ (298.15 K),
*T* is the temperature of the dissolution system. $-\Delta_{sol}H/R$ is
a constant refers to the enthalpy of dissolution which does not change
much with temperature. The $H^{cc}_{\theta}$ and $-\Delta_{sol}H/R$ of
CO2 are 0.83 and 2400 K, respectively.

**Henry's law** can be used for modeling the solubility of CO~2~ in
aqueous solution for pressures up to 100 MPa
`(Carrol and Mather, 1992)`. Therefore, the pressure posed little
influence on the calculation.

On the other hand,the fraction of the total dissolved as CO~2~(aq)
changes with pH, while the *C~aq~* is controlled by the Henry's law.

CO~2~(aq) + H~2~O ⇋ H~2~CO~3~

H~2~CO~3~ + H~2~O ⇋ H~3~O^+^ + HCO~3~^−^ *K~a1~ = 4.45 10^-7^* at 25
^o^C

HCO~3~^−^ + H~2~O ⇋ H~3~O^+^ + CO~2~^−^ *K~a2~ = 4.67 10^-11^* at 25
^o^C

Therefore, the total CO~2~(aq)/CO~2~(aq) = 1 + *K~a1~*/[H^+^] +
*K~a1~*\*\*K~a2~\*/[H^+^]^2^

The *K~a~* values are temperature dependent. According to Van't Hoff
equation:

$$dln(K_a)/dT = \Delta H^{\theta}/RT^2$$
$$lnK_2/K_1 = -\Delta H^{\theta}/R (1/T_2 - 1/T_1)$$

Based on these theories, we could calculate the ratios between CO~2~ in
aqueous solution and in gas phase.

```{r eval=FALSE}
CO2_dissolution <- function (T, pH) { #input the T in degree C
  ka1 <- 4.45*10^-7
  ka2 <- 4.67*10^-11
  Hi <- 10^(-pH)
  vant1 <- 2400*(1/(T+273.15)-1/298.15)
  vant2 <- exp(vant1)
  CO2_pH <- 1+ka1*vant2/Hi+ka1*ka2*vant2*vant2/(Hi^2)
  CO2_aq <- 0.83*vant2
  total_CO2 <- CO2_pH*CO2_aq
  return (total_CO2) #ratios between CO~2~ in aqueous solution and in gas phase.
}

```

## d18O-temperature

```{r}
calcite_d18O_temperature <- function (T) { #T in degree C
  T <- T+273.15
  2.78*10^6/T^2-2.89 #Friedman and O'Neil, 1977, Kim and O'Neil, 1997
}

d18O_temperature_calcite <- function (x) { #input the d18O of calcite veins, return the precipitation temperture in degree C
    temp <- (x+2.89)/2.78
    temperature <- sqrt(10^6/temp)-273.15 #Friedman and O'Neil, 1977, Kim and O'Neil, 1997
    return (temperature)
  }


d18O_temperature_aragonite <- function (x) { #Kim et al., 2007
  17.88*10^3/(x+31.14)-273.15 #input d18O of aragonite, return temperature
}

dolomite_d18O_temperature <- function (T) { #T in degree C
  T <- T+273.15
  3.14*10^6/T^2-3.14 #Horita, 2014
}



```

## REE normalization

```{r}

REE_normalization <- function (data, method) { #method should be CI or PAAS
  data_temp <- data %>% select(-La,-Ce,-Pr,-Nd,-Sm,-Eu,
                               -Gd,-Tb,-Dy,-Y,-Ho,-Er,
                               -Tm,-Yb,-Lu)
  if (method=="CI") { #CI: CI chondrite
    data$La=data$La/0.237
    data$Ce=data$Ce/0.613
    data$Pr=data$Pr/0.0928
    data$Nd=data$Nd/0.457
    data$Sm=data$Sm/0.148
    data$Eu=data$Eu/0.0563
    data$Gd=data$Gd/0.199
    data$Tb=data$Tb/0.0361
    data$Dy=data$Dy/0.246
    data$Y=data$Y/1.57
    data$Ho=data$Ho/0.0546
    data$Er=data$Er/0.16
    data$Tm=data$Tm/0.0247
    data$Yb=data$Yb/0.161
    data$Lu=data$Lu/0.0246
    data_temp2 <- data.frame(La=data$La,Ce=data$Ce,Pr=data$Pr,Nd=data$Nd,
                      Sm=data$Sm,Eu=data$Eu,Gd=data$Gd,Tb=data$Tb,
                      Dy=data$Dy,Y=data$Y,Ho=data$Ho,Er=data$Er,
                      Tm=data$Tm,Yb=data$Yb,Lu=data$Lu)
  data_normalized <- bind_cols(data_temp,data_temp2)
  return(data_normalized)
    
  } else if (method=="PAAS") { #PAAS
    data$La=data$La/38.2
    data$Ce=data$Ce/79.6
    data$Pr=data$Pr/8.83
    data$Nd=data$Nd/33.9
    data$Sm=data$Sm/5.55
    data$Eu=data$Eu/1.08
    data$Gd=data$Gd/4.66
    data$Tb=data$Tb/0.774
    data$Dy=data$Dy/4.68
    data$Y=data$Y/27
    data$Ho=data$Ho/0.991
    data$Er=data$Er/2.85
    data$Tm=data$Tm/0.405
    data$Yb=data$Yb/2.82
    data$Lu=data$Lu/0.433
    data_temp2 <- data.frame(La=data$La,Ce=data$Ce,Pr=data$Pr,Nd=data$Nd,
                      Sm=data$Sm,Eu=data$Eu,Gd=data$Gd,Tb=data$Tb,
                      Dy=data$Dy,Y=data$Y,Ho=data$Ho,Er=data$Er,
                      Tm=data$Tm,Yb=data$Yb,Lu=data$Lu)
  data_normalized <- bind_cols(data_temp,data_temp2)
  return(data_normalized)
    
  } else return("Please input the correct normalization method, CI or PAAS")
}

```
